---
title: 'PSY 710: Assignment 3'
author: "Tinashe M. Tapera"
output:
  pdf_document: default
  html_notebook: default
---

> Use the "default" data set to run a logistic regression model with Student as the only independent variable to predict whether customers are likely to default on their credit card or not.

```{r}
setwd("/Users/mac/Dropbox/Drexel\ U/Psychology/PSY\ 710/PSY710_Assignments")
library(foreign)
default = read.spss("default.sav", to.data.frame =  TRUE)
default$default = ifelse(default$default == "no", 0,1)
default$student = ifelse(default$student == "No", 0,1)
head(default)
model = glm(default ~ student, data = default, family = "binomial")
summary(model)
```

> a) Based on the model output, what is the odds of a customer who is likely to default given that he or she is a student?

Using the probability formula given by the logit function:

$$P(default) = \frac{e^{b_0 + b_1X}} {1+e^{b_0 + b_1X}}$$

The probability that someone defaults (1), given that they are a student (1) is:
```{r}
P_default = exp(as.numeric(model$coefficients[1])+
                as.numeric(model$coefficients[2])*1)/(
                1+exp(as.numeric(model$coefficients[1])+
                as.numeric(model$coefficients[2])*1))
P_default
```

The probability that they don't default, given that they are a student:
```{r}
p_prime = 1-P_default
p_prime
```

Therefore the odds of someone defaulting given that they are a student is given by:
```{r}
P_default/p_prime
```

> b) What is the odds of a customer who is likely to default given that the customer is not a student?

The probability that someone defaults (1), given that they are not a student (0) is:
```{r}
P_default = exp(as.numeric(model$coefficients[1])+
                as.numeric(model$coefficients[2])*0)/(
                1+exp(as.numeric(model$coefficients[1])+
                as.numeric(model$coefficients[2])*0))
P_default
```

The probability that they don't default, given that they are not a student:
```{r}
p_prime = 1-P_default
p_prime
```

Therefore the odds of someone defaulting given that they are not a student is given by:
```{r}
P_default/p_prime
```

> c) What is the odds ratio and how to interpret it?

The odds ratio is:
```{r}
0.04508342/0.03007299
```

This means that the odds of a person defaulting is 1.499 times higher if they are a student, than if not.